---
title: "07.2 - Murine models of NB-FOXR2: validation"
author: "Selin Jessa [[selin.jessa@mail.mcgill.ca](mailto:selin.jessa@mail.mcgill.ca)] and Bhavyaa Chandarana [[bhavyaa.chandarana@mail.mcgill.ca](mailto:bhavyaa.chandarana@mail.mcgill.ca)]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    self_contained: yes
    keep_md: yes
    code_folding: show
    theme: flatly
    css: ../include/style.css
    toc: yes
    toc_depth: 4
    number_sections: true
    df_print: paged
    includes:
      before_body: ../include/header.html
      after_body:  ../include/footer.html
---

<!-- FRONT MATTER, insert configuration info -->

```{r header, echo = FALSE, results = 'asis', warning = FALSE}

# Index of the document
# ...determines name of the subfolder of `outputs` and `figures`
doc_id <- "07.2"

suppressMessages(library(here))

# Knit child document with header
res <- knitr::knit_child(here("include", "header.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF FRONT MATTER -->


# Overview

In this document, we analyze murine models generated through _in utero_ electroporation
of constructs with Foxr2, or Foxr2 and p53 LOF, into the mouse ganglionic
eminences. 

Here, we ask whether the IUE mouse models (profiled by 10X Multiome) are good models 
of human NB-FOXR2 tumors. Are they more similar to NB-FOXR2 than to other pediatric 
brain tumors?

# Libraries & functions

```{r libs, warning = FALSE, message = FALSE, echo = TRUE, cache = FALSE}

library(here) 
library(magrittr)
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(stringr)
library(glue)
library(purrr)
library(ggplot2)
library(cowplot)
library(Seurat)
library(Signac)
library(ComplexHeatmap)
library(tibble)
library(ggridges)
library(GSVA)
library(ggVennDiagram)
library(ggrepel)
library(plotly)

source(here("include/style.R"))
source(here("code/functions/scRNAseq.R"))
source(here("code/functions/RNAseq.R"))
source(here("code/functions/ssGSEA.R"))

ggplot2::theme_set(theme_min())

square_theme <- theme(aspect.ratio = 1)
large_text <- theme(text = element_text(size = 15))

conflicted::conflicts_prefer(base::setdiff)
conflicted::conflicts_prefer(dplyr::filter)

conflicted::conflicts_prefer(graphics::layout)

```

Functions to convert genes from hg to mm. Functions were adapted from in-house bulk pipeline by Steven Hébert.

<details>

```{r hg_to_mm_function, class.source = 'fold.hide'}

mouse_human_genes <- data.table::fread("/project/kleinman/steven.hebert/from_narval/2024/2023-05-16-FOXR2paper/2024-01-29-revision/2024-03-06-human_NB-FOXR2_signatures_bulk/signature_enrichment_Compendium/input/mm10_hg19_ensembl_symbol_raptor_unique.tsv")

# Convert list of genes from human to mouse
# Return mouse genes
hg_to_mm_pipeline <- function(hg_genes, ref) {

  # To keep the order of the genes supplied, we need to join using the hg_genes vector
  hg_ensembl = map_chr(str_split(hg_genes, ":"),1) %>% as.data.frame()
  colnames(hg_ensembl) = "hg_ensembl_gene_id"
  mm_ID = hg_ensembl %>%
    inner_join(.,ref) %>%
    mutate(mm_ID = paste(mm_ensembl_gene_id, mm_symbol_id, sep = ":")) %>%
    pull(mm_ID)

  return(mm_ID)
}

# Convert list of genes from mouse to human
# Return human genes
mm_to_hg_pipeline <- function(mm_genes, ref) {

  # To keep the order of the genes supplied, we need to join using the mm_genes vector
  mm_ensembl = map_chr(str_split(mm_genes, ":"),1) %>% as.data.frame()
  colnames(mm_ensembl) = "mm_ensembl_gene_id"
  hg_ID = mm_ensembl %>%
    inner_join(.,ref) %>%
    mutate(hg_ID = paste(hg_ensembl_gene_id, hg_symbol_id, sep = ":")) %>%
    pull(hg_ID)

  return(hg_ID)
}

# Convert genes from human to mouse
# Returns df of equivalent human and mouse genes (dataframe)
hg_to_mm_pipeline_df <- function(hg_genes, ref){

  # To keep the order of the genes supplied, we need to join using the hg_genes vector
  hg_ensembl = map_chr(str_split(hg_genes, ":"),1) %>% as.data.frame()
  colnames(hg_ensembl) = "hg_ensembl_gene_id"
  mm_ID = hg_ensembl %>%
    inner_join(.,ref) %>%
    mutate(mm_ID = paste(mm_ensembl_gene_id, mm_symbol_id, sep = ":"),
           hg_ID = paste(hg_ensembl_gene_id, hg_symbol_id, sep = ":"))

 }

```

</details>

# Load mouse model single-cell data

We have two genotypes:

1. Foxr2 overexpression (OE) with p53LOF (sample `Foxr2_p53_r1` aka `AN24377`)
2. Foxr2 overexpression (OE) alone (samples `Foxr2_r1` aka `M7238`, and `Foxr2_r2` aka `AN22476`)

Each sample comes from a separate mouse, hence these are biological, not technical replicates.

```{r load_samples, cache = FALSE}

mouse_samples <- c("Foxr2_p53_r1" = "AN24377",
                   "Foxr2_r1"     = "M7238",
                   "Foxr2_r2"     = "AN22476")

# load all three objects into a list
seurat_mouse <- map(mouse_samples, ~ get(load(here(glue("data/singlecell/pipeline_scMultiome_mm/{.x}/seurat.Rda")))))
names(seurat_mouse) <- names(mouse_samples)

# For now, we will not use the ATAC, so let's make a "light" version of the objects
seurat_mouse <- map(seurat_mouse,
                    function(seurat){
                        
                            seurat[['ATAC']] <- NULL
                            seurat[['peaks']] <- NULL
                            seurat[['promoters']] <- NULL
                        
                        return(seurat)
                    })

load(here("output/07.1/seurat_mouse_metadata.Rda"))
seurat_mouse <- map2(seurat_mouse, seurat_mouse_metadata,
                     function(seurat, meta) {
                         seurat@meta.data <- meta %>% 
                             column_to_rownames("Cell")
                         return(seurat)
                     })


```

# Comparison with human tumors

We will compare single-nuclei transcriptomic datasets of mouse models to bulk 
transcriptomic datasets of human pediatric brain tumors (PBT). 

We will use the following comparators for pairwise differential expression:

* **DIPG-H3K27M** (glial tumor)
* **MB-WNT** (neuronal tumor)
* **ETMR** (neuronal tumor)
* **NB-FOXR2** 

For the NB-FOXR2, we observe batch effect between the in-house samples 
and samples from Korshunov publication, so we restricted to in-house samples
for this differential expression.

For the DIPG-H3K27M group, we removed the following three samples with 
low tumor purity (see ssGSEA tumor signature scoring "Top 10 signatures per sample"
in R markdown document `06.1`):

* P-6254_S-8499
* P-5092_S-7792
* P-6253_S-8498

Note: in each case the NB-FOXR2 "side" of the DE corresponds to the negative 
log fold change values, so I will filter signatures with this in mind.

We will adapt signature filtering methods used in [McNicholas et al. *Cancer Discovery* 2023](https://pubmed.ncbi.nlm.nih.gov/37011011/) to create main Figure 3B.

## Create consensus human tumor signatures

Generate consensus signatures (across DE comparisons) for selected tumors to be
compared with NB-FOXR2.

First define a function to extract signatures from each DESeq result:

<details>

```{r function-extract-sigs, message = F, warning = FALSE}

# Function to extract up- and down-regulated DE signatures from each DESeq result file.
# - Set slice = NULL to take all DE genes.
# - Group 1 of DESeq comparison (level3 pipeline) is negative fold change.
# - Variable tumor_type_comparison is string in format "Tumor1vsTumor2".
extract_sigs_all <- function(sigs_path,
                             out_path,
                             tumor_type_comparison,
                             convert_hg_to_mm = FALSE,
                             slice = 100){
    
    df <- data.table::fread(glue("{sigs_path}/{tumor_type_comparison}.tsv"))
    head(df %>% arrange(stat))
    head(df %>% arrange(desc(stat)))
    
    tumor_type_1 <- strsplit(tumor_type_comparison,
                            split = "vs")[[1]][1]
    tumor_type_2 <- strsplit(tumor_type_comparison,
                            split = "vs")[[1]][2]
    
    sigs <- c()
    sigs[[glue("{tumor_type_1}_vs_{tumor_type_2}")]] <- df %>% separate(ID, 
                                   into = c("ENSG", "symbol"), 
                                   sep = ":", remove = F) %>% 
        filter(padj < 0.05 & baseMean > 100 & abs(log2FoldChange) > 1) %>%
        filter(log2FoldChange < 0) %>% # Negative log fold change
        filter(!grepl("^MT-", symbol)) %>% 
        arrange(stat) %>% 
        pull(ID)
    
    sigs[[glue("{tumor_type_2}_vs_{tumor_type_1}")]] <- df %>% separate(ID, 
                                   into = c("ENSG", "symbol"), 
                                   sep = ":", remove = F) %>% 
        filter(padj < 0.05 & baseMean > 100 & abs(log2FoldChange) > 1) %>%
        filter(log2FoldChange > 0) %>% # Positive log fold change
        filter(!grepl("^MT-", symbol)) %>% 
        arrange(desc(stat)) %>% 
        pull(ID)
    
    if(convert_hg_to_mm){
        
        message(glue("@ {tumor_type_1} hg to mm"))
        sigs[[glue("{tumor_type_1}_vs_{tumor_type_2}")]] <- sigs[[glue("{tumor_type_1}_vs_{tumor_type_2}")]] %>% 
            hg_to_mm_pipeline(mouse_human_genes)
        
        message(glue("@ {tumor_type_2} hg to mm"))
        sigs[[glue("{tumor_type_2}_vs_{tumor_type_1}")]] <- sigs[[glue("{tumor_type_2}_vs_{tumor_type_1}")]] %>% 
            hg_to_mm_pipeline(mouse_human_genes)

        
    } 
    
    if(!is.null(slice)){
        
        message(glue("@ {tumor_type_1} top {slice}"))
        sigs[[glue("{tumor_type_1}_vs_{tumor_type_2}")]] <- sigs[[glue("{tumor_type_1}_vs_{tumor_type_2}")]] %>% 
            .[1:slice]
        
        message(glue("@ {tumor_type_2} top {slice}"))
        sigs[[glue("{tumor_type_2}_vs_{tumor_type_1}")]] <- sigs[[glue("{tumor_type_2}_vs_{tumor_type_1}")]] %>% 
            .[1:slice]
        
    }
    
    save(sigs, file = glue("{out_path}/{tumor_type_1}_vs_{tumor_type_2}_signatures.Rda"))
    
    data.table::fwrite(sigs[[glue("{tumor_type_1}_vs_{tumor_type_2}")]] %>% as.data.frame(),
                       file = glue("{out_path}/{tumor_type_1}_vs_{tumor_type_2}_signatures{slice}.tsv", .null = ""),
                       col.names = F)
    data.table::fwrite(sigs[[glue("{tumor_type_2}_vs_{tumor_type_1}")]] %>% as.data.frame(),
                       file = glue("{out_path}/{tumor_type_2}_vs_{tumor_type_1}_signatures{slice}.tsv", .null = ""),
                       col.names = F)
    
    return(sigs)
}
```

</details>

1. Extract 2 lists of differentially expressed genes (upregulated & downregulated) per 
DESeq comparison, and save.

```{r comparator-sigs, message = F, warning = FALSE}

# Get the list of DESeq comparisons from the level3 output folder
sigs_path_comparators <- here("data/RNAseq/pipeline_l3/2024-04-human_comparator_tumor_signatures_rem_low_purity_DIPG/diff/Ensembl.ensGene.exon")
comparisons <- list.files(sigs_path_comparators) %>% 
    .[!grepl(x = ., pattern = ".QC")] %>% 
    sub(pattern = ".tsv", replacement = "")
comparisons <- comparisons[!grepl(comparisons, pattern = "HGG-IDH")]

# Produces a list of paired signatures (up and down from each DESeq result) n=100 genes
sigs_list_comparators_100 <- purrr::map(comparisons,
        ~ extract_sigs_all(sigs_path = sigs_path_comparators,
                     tumor_type_comparison = .x, 
                     convert_hg_to_mm = TRUE,
                     out_path = glue("{out}/signatures_v2_comparators")))
names(sigs_list_comparators_100) <- comparisons
save(sigs_list_comparators_100, 
    file = glue("{out}/sigs_list_comparators_100.Rda"))

# Produces a list of paired signatures (up and down from each DESeq result) ALL DE genes
sigs_list_comparators <- purrr::map(comparisons,
        ~ extract_sigs_all(sigs_path = sigs_path_comparators,
                     tumor_type_comparison = .x, 
                     convert_hg_to_mm = TRUE,
                     out_path = glue("{out}/signatures_v2_comparators"),
                     slice = NULL))
names(sigs_list_comparators) <- comparisons
save(sigs_list_comparators, 
    file = glue("{out}/sigs_list_comparators.Rda"))

```

2. Get consensus signatures for each of the tumors.

```{r consensus-comparators}

comparator_tumor_types <- c("DIPG-H3K27M",
                            "ETMR",
                            "MB-WNT",
                            "NB-FOXR2")

# 1. Un-pair the signatures, so that they are all individual elements in one list
# Get first list of each pair and concatenate:
sigs_list_comparators_1 <- map(comparisons,
                             ~ sigs_list_comparators[.x][[1]][1]) %>% Reduce(f = c)
# Get second list of each pair, concatenate, and then merge with list above
sigs_list_comparators_2 <- map(comparisons,
                             ~ sigs_list_comparators[.x][[1]][2]) %>% Reduce(f = c)
sigs_list_comparators_unpair <- c(sigs_list_comparators_1, sigs_list_comparators_2)

# 2. Take the consensus with intersect for each of those comparator tumor types listed
consensus_all_DE_comparators <- map(comparator_tumor_types,
                                    ~ sigs_list_comparators_unpair[grepl(x = names(sigs_list_comparators_unpair), 
                                                                  pattern = glue("^{.x}"))] %>% 
                                        Reduce(f = intersect))
names(consensus_all_DE_comparators) <- comparator_tumor_types
save(consensus_all_DE_comparators, 
     file = glue("{out}/consensus_all_DE_comparators.Rda"))

map(consensus_all_DE_comparators, length)

# 3. Get human version of comparator signatures 
consensus_all_DE_comparators_hg <- map(consensus_all_DE_comparators,
                                       ~ mm_to_hg_pipeline(mm_genes = .x, ref = mouse_human_genes))
# Check no genes are lost during comparison back to human
map(consensus_all_DE_comparators_hg, length)
save(consensus_all_DE_comparators_hg, 
     file = glue("{out}/consensus_all_DE_comparators_hg.Rda"))
```

3. Cut each comparator signature to 100 genes, based on average stat value across 
DESeq comparisons with that tumor type.

Define function to cut top 100 genes by stat value:

<details>

```{r}

# Function to cut top 100 genes per signature
cut_100_stat <- function(tumor_type_comparison_list, 
                         tumor_type,
                         gene_list,
                         sigs_path){
    
    comparison_results <- data.frame(ID = paste0(mouse_human_genes$hg_ensembl_gene_id, ":", mouse_human_genes$hg_symbol_id))
    
    for(tumor_type_comparison in tumor_type_comparison_list){
        
        tumor_type_1 <- strsplit(tumor_type_comparison,
                            split = "vs")[[1]][1]
        tumor_type_2 <- strsplit(tumor_type_comparison,
                            split = "vs")[[1]][2]
    
        if(tumor_type %in% c(tumor_type_1, tumor_type_2)){
            
            df <- data.table::fread(glue("{sigs_path}/{tumor_type_comparison}.tsv"))
        
            if(tumor_type == tumor_type_1){ # Negative log fold change & stat value
                
                df <- df %>% 
                    select(ID, stat) %>% 
                    mutate(stat = stat*-1) # Flip sign on values
                
                
            } else if(tumor_type == tumor_type_2) { # Positive log fold change & stat value
                
                df <- df %>% 
                    select(ID, stat) # Take values as-is
                
            }
            
            colnames(df)[which(colnames(df) == "stat")] <- tumor_type_comparison
            
            comparison_results <- left_join(comparison_results, df, by = "ID")
        
        }
    }
    
    comparison_results$mean_stat <- rowMeans(comparison_results[,-1])
    
    top_100 <- comparison_results %>% 
        filter(ID %in% gene_list) %>% 
        arrange(desc(mean_stat)) %>% 
        .$ID %>% 
        .[1:100]
    
    return(top_100)
    
}

```

</details>

Get top 100 genes:

```{r}

# Run function
consensus_all_DE_comparators_hg_100 <- imap(consensus_all_DE_comparators_hg,
                                           ~ cut_100_stat(gene_list = .x,
                                                          tumor_type = .y,
                                                          tumor_type_comparison_list = comparisons,
                                                          sigs_path = sigs_path_comparators))
map(consensus_all_DE_comparators_hg_100, length)
save(consensus_all_DE_comparators_hg_100, 
     file = glue("{out}/consensus_all_DE_comparators_hg_100.Rda"))

# Get mouse version as well
consensus_all_DE_comparators_mm_100 <- map(consensus_all_DE_comparators_hg_100,
                                       ~ hg_to_mm_pipeline(hg_genes = .x, ref = mouse_human_genes))
map(consensus_all_DE_comparators_mm_100, length)
save(consensus_all_DE_comparators_mm_100, 
     file = glue("{out}/consensus_all_DE_comparators_mm_100.Rda"))

# Set up list of 6 signatures (NB-FOXR2 consensus sig from 8 signatures + 5 comparator signatures)
sigs_list_ssgsea <- map(consensus_all_DE_comparators_mm_100,
                        ~ map_chr(str_split(.x, ":"),2))
names(sigs_list_ssgsea) <- names(consensus_all_DE_comparators_mm_100)

save(sigs_list_ssgsea,
     file = glue("{out}/sigs_list_ssgsea_mm.Rda"))

# Do the same for human
sigs_list_ssgsea_hg <- map(consensus_all_DE_comparators_hg_100,
                        ~ map_chr(str_split(.x, ":"),2))
names(sigs_list_ssgsea_hg) <- names(consensus_all_DE_comparators_hg_100)

save(sigs_list_ssgsea_hg,
     file = glue("{out}/sigs_list_ssgsea_hg.Rda"))

```

## NB-FOXR2 Signature volcano plots

Here, we produce volcano plots of NB-FOXR2 vs. each other tumor type to illustrate
the method of creating consensus signatures for the publication.

We color here in red the 100 genes that appear in the final consensus signature for
NB-FOXR2 across comparisons.

Define function to plot volcano plots with highlighted genes 
(modified from a function by Alva Annett).

<details>

```{r volcano-plot-function}

volcano_gene_set <- function(df, log2fc = 1, pval=0.05, bm = 100, genes, name = '', labeled_genes = NULL){
  
  df <- df %>% 
    filter(baseMean > bm)
  
  p <- df %>%
    mutate(col = ifelse(SYM %in% genes, 'y', NA)) %>%
    arrange(!is.na(col), col) %>% 
    ggplot(aes(x=-log2FoldChange, y=-log10(padj), col=col, label = SYM)) +
    geom_point(size=2, alpha=0.7) +
    scale_color_manual(values=c(y = "red"), 
                       na.value = "grey40") +
    geom_vline(xintercept=c(-log2fc, log2fc), col="grey50", linetype='dotted') +
    geom_hline(yintercept=-log10(pval), col="grey50", linetype='dotted') +
    theme_bw() +
    #theme(text=element_text(size=13), legend.position = 'none') +
    xlab("log2 Fold Change") +
    ylab("-log10 Adjusted p-value") +
    theme(text = element_text(size = 15),
          axis.line = element_line(size = 0.5),
          panel.border = element_blank(),
          panel.grid = element_blank()) +
    notebook_theme +
    ggtitle(name) 

    
    if(!is.null(labeled_genes)){
        p <- p + geom_label_repel(data = df %>% filter(SYM %in% labeled_genes),
                                 fill = "white",
                                 show.legend = F, box.padding = 0.5, 
                                 max.overlaps = Inf, color = "black") +
            labs(title = "min.segment.length = 0") +
            notebook_theme +
            ggtitle(name)
    }
  
  return(ggrastr::rasterise(p))
}

notebook_theme <- theme(aspect.ratio = 1,
                       text = element_text(size = 15), 
                       panel.border = element_rect(colour = "grey85", fill=NA, size=1),
                       axis.ticks = element_line(colour = "grey85"),
                       panel.grid.major = element_blank(), 
                       panel.grid.minor = element_blank(),
                       panel.background = element_blank(),
                       axis.line = element_blank(), 
                       legend.position = "none",
                       strip.background = element_blank())

volcano_gene_set_label <- function(df, log2fc = 1, pval=0.05, bm = 100, genes, name = '', labeled_genes = NULL){
  
  df <- df %>% 
    filter(baseMean > bm)
  
  p <- df %>%
    mutate(col = ifelse(SYM %in% genes, 'y', NA)) %>%
    arrange(!is.na(col), col) %>% 
    ggplot(aes(x=-log2FoldChange, y=-log10(padj), col=col, label = SYM)) +
    geom_point(size=2, alpha=0.7) +
    scale_color_manual(values=c(y = "red"), 
                       na.value = "grey40") +
    theme_bw() +
    xlab("log2 Fold Change") +
    ylab("-log10 Adjusted p-value") +
    notebook_theme +
    theme(aspect.ratio = 1/0.75) +
    ggtitle(name) 

    
    if(!is.null(labeled_genes)){
        p <- p + geom_text_repel(data = df %>% filter(SYM %in% labeled_genes),
                                 min.segment.length = 0, box.padding = 0.5,
                                 max.overlaps = Inf, color = "black") +
            notebook_theme +
            theme(aspect.ratio = 1/0.75) +
            ggtitle(name)
    }
  
  return(ggrastr::rasterise(p))
}



```

</details>

Get the required data from DESeq2 results to produce volcano plots:

```{r volcano-nb-foxr2-sig}

# Get DESeq results for comparisons including NB-FOXR2
sigs_path_comparators <- here("data/RNAseq/pipeline_l3/2024-04-human_comparator_tumor_signatures_rem_low_purity_DIPG/diff/Ensembl.ensGene.exon")
comparisons <- list.files(sigs_path_comparators) %>% 
    .[!grepl(x = ., pattern = ".QC")] %>% 
    sub(pattern = ".tsv", replacement = "")
nbfoxr2_comparisons <- comparisons[grepl(comparisons, pattern = "NB-FOXR2")] %>% 
    .[!grepl(., pattern = "HGG-IDH")]

deseq_dfs_nb_foxr2 <- map(nbfoxr2_comparisons,
                          ~ data.table::fread(glue("{sigs_path_comparators}/{.x}.tsv")) %>% 
                            tidyr::separate(col = "ID", into = c("ENS", "SYM"), sep = ":"))
names(deseq_dfs_nb_foxr2) <- nbfoxr2_comparisons

# Get the genes in the NB-FOXR2 signature
sym_nb_foxr2 <- map_chr(str_split(consensus_all_DE_comparators_hg_100[["NB-FOXR2"]], ":"),2)

```

### Interactive volcano

Plot an interactive (plotly) version of volcano plot to easily check genes
appearing in the comparison.

```{r volcano-interactive}

volcano_gene_set(deseq_dfs_nb_foxr2[[1]],
                 genes = sym_nb_foxr2,
                 name = names(deseq_dfs_nb_foxr2)[1]) %>% ggplotly()

volcano_gene_set(deseq_dfs_nb_foxr2[[2]],
                 genes = sym_nb_foxr2,
                 name = names(deseq_dfs_nb_foxr2)[2]) %>% ggplotly()

volcano_gene_set(deseq_dfs_nb_foxr2[[3]],
                 genes = sym_nb_foxr2,
                 name = names(deseq_dfs_nb_foxr2)[3]) %>% ggplotly()

```


### Labelled volcano

Label specific genes in each volcano for publication:

```{r volcano-labeled}

volcano_gene_set_label(deseq_dfs_nb_foxr2[["NB-FOXR2vsDIPG-H3K27M"]],
                 genes = sym_nb_foxr2,
                 name = "NB-FOXR2vsDIPG-H3K27M",
                 labeled_genes = c("NKX2-1", "LHX6", "FOXG1",
                                   "GFAP", "AQP4", "MBP", "MOG"))


volcano_gene_set_label(deseq_dfs_nb_foxr2[["NB-FOXR2vsETMR"]],
                 genes = sym_nb_foxr2,
                 name = "NB-FOXR2vsETMR",
                 labeled_genes = c("GPD1", "FAM163A", "FERMT1",
                                   "LIN28A"))


volcano_gene_set_label(deseq_dfs_nb_foxr2[["NB-FOXR2vsMB-WNT"]],
                 genes = sym_nb_foxr2,
                 name = "NB-FOXR2vsMB-WNT",
                 labeled_genes = c("NKX2-1", "ETS1", "FAM163A", "DNM3",
                                   "ZIC5", "NKD1", "ZIC1", "ZIC3"))


```

## Signature QC: ssGSEA in human tumors

Here, we test the quality of the generated consensus signatures by scoring them
in human tumors, both bulk and single-cell.

### Bulk

1. Load NB-FOXR2 bulk tumor counts matrices (raw) for the full tumor cohort.

2. Subset the tumor cohort into their individual tumor types (and split NB-FOXR2 
into Korshunov and non-Korshunov/in-house).

```{r load-bulk}

# Load bulk metadata
meta <- read_tsv(here("output/00/metadata_patients_NGS.tsv"))
meta_bulk <- meta %>% filter(RNAseq == "Y")

# Get sample names for each tumor type
# Alter the names with gsub to match the sample names found in counts matrix column names 
# (done manually by comparing the sample names)

NBFOXR2_bulk_sample_names_Korshunov <- meta_bulk %>% 
    filter(Group == "NB-FOXR2") %>% 
    filter(grepl(x = .$Source, pattern = "Korshunov")) %>% 
    .$ID %>% 
    gsub(pattern = "^", replacement = "X") %>% 
    gsub(pattern = "$", replacement = ".0")

NBFOXR2_bulk_sample_names_noKorshunov <- meta_bulk %>% 
    filter(Group == "NB-FOXR2") %>% 
    filter(!grepl(x = .$Source, pattern = "Korshunov")) %>% 
    .$ID %>% 
    gsub(pattern = "P-", replacement = "P.") %>% 
    gsub(pattern = "S-", replacement = "S.") %>% 
    gsub(pattern = "-S", replacement = "_S") %>% 
    gsub(pattern = "-P", replacement = "_P")  %>% 
    gsub(pattern = "MDT-AP", replacement = "MDT.AP") %>% 
    gsub(pattern = "15-", replacement = "15.")

MBWNT_bulk_sample_names <- meta_bulk %>% 
    filter(Group == "MB-WNT") %>% 
    .$ID %>% 
    gsub(pattern = "P-", replacement = "P.") %>% 
    gsub(pattern = "S-", replacement = "S.") %>% 
    gsub(pattern = "-S", replacement = "_S") %>% 
    gsub(pattern = "-P", replacement = "_P")  %>% 
    gsub(pattern = "MDT-AP", replacement = "MDT.AP") %>% 
    gsub(pattern = "15-", replacement = "15.")

ETMR_bulk_sample_names <- meta_bulk %>% 
    filter(Group == "ETMR") %>% 
    .$ID %>% 
    gsub(pattern = "P-", replacement = "P.") %>% 
    gsub(pattern = "S-", replacement = "S.") %>% 
    gsub(pattern = "-S", replacement = "_S") %>% 
    gsub(pattern = "-P", replacement = "_P")  %>% 
    gsub(pattern = "MDT-AP", replacement = "MDT.AP") %>% 
    gsub(pattern = "15-", replacement = "15.")

DIPG_bulk_sample_names <- meta_bulk %>% 
    filter(Group == "DIPG-H3K27M") %>% 
    .$ID %>% 
    gsub(pattern = "P-", replacement = "P.") %>% 
    gsub(pattern = "S-", replacement = "S.") %>% 
    gsub(pattern = "-S", replacement = "_S") %>% 
    gsub(pattern = "-P", replacement = "_P")  %>% 
    gsub(pattern = "MDT-AP", replacement = "MDT.AP") %>% 
    gsub(pattern = "15-", replacement = "15.") %>% 
    gsub(pattern = "00-", replacement = "X00.")

sample_names <- list("NBFOXR2_bulk_sample_names_noKorshunov" = NBFOXR2_bulk_sample_names_noKorshunov,
                    "NBFOXR2_bulk_sample_names_Korshunov" = NBFOXR2_bulk_sample_names_Korshunov,
                    "MBWNT_bulk_sample_names" = MBWNT_bulk_sample_names,
                    "ETMR_bulk_sample_names" = ETMR_bulk_sample_names,
                    "DIPG_bulk_sample_names" = DIPG_bulk_sample_names)

# Load bulk counts matrix and subset based on sample name lists set up above
counts_all_tumors <- read.table("/project/kleinman/steven.hebert/from_narval/2024/2023-05-16-FOXR2paper/2024-01-29-revision/2024-01-31-PCA-pipelineGenes/level3/hg19/all_human_tumors_v0/counts/Ensembl.ensGene.exon.norm.tsv.gz", sep = "\t", row.names = 1, header = T) %>% as.matrix()

counts_dfs <- map(sample_names,
                 ~ counts_all_tumors %>% 
                     as.data.frame() %>% 
                    select(all_of(.x)) %>% 
                    as.matrix())

names(counts_dfs) <- c("NB-FOXR2 in-house",
                      "NB-FOXR2 Korshunov",
                      "MB-WNT",
                      "ETMR",
                      "DIPG")

save(counts_dfs, file = glue("{out}/counts_dfs.Rda"))

```

Run GSVA in each of the bulk tumor counts matrices.

```{r run-gsva-bulk, message = FALSE, warning = FALSE, results = FALSE}

sigs_list_ens_sym_hg <- consensus_all_DE_comparators_hg_100
save(sigs_list_ens_sym_hg, file = glue("{out}/sigs_list_ens_sym_hg.Rda"))

ssGSEA_scores_bulk <- map(counts_dfs,
                          ~ GSVA::gsva(.x,
                                    sigs_list_ens_sym_hg,
                                    method="ssgsea",
                                    ssgsea.norm = F,
                                    tau = 0.25))
names(ssGSEA_scores_bulk) <- names(counts_dfs)

save(ssGSEA_scores_bulk, file = glue("{out}/ssGSEA_scores_bulk.Rda"))

```

Plot scores in the bulk samples of each tumor type.

```{r gsva-plot-bulk-per-tumor, fig.width = 20, fig.height=12, cache = F}

load(file = glue("{out}/ssGSEA_scores_bulk.Rda"))

plot_scores_across_signatures <- function(ssGSEA_score_df,
                                          tumor_type){
    
    ssGSEA_score_df %>% t() %>% 
          as.data.frame() %>%
          rownames_to_column(var = "sample") %>%
          pivot_longer(cols = -sample, 
                       values_to = "ssGSEAscore",
                       names_to = "signature") %>%
            ggplot(aes(x = factor(signature),
                        y = ssGSEAscore,
                        fill = signature)) +
                geom_violin() +
                geom_boxplot(width=0.1, fill = "white") +
                scale_fill_manual(values = palette_groups) +
                theme_classic() +
                ggtitle(glue("{tumor_type}: scores per signature")) +
                xlab("Signature") +
                theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
                        plot.title = element_text(hjust = 0.5, size = 20)) + 
                no_legend()
    
}

imap(ssGSEA_scores_bulk,
      ~ plot_scores_across_signatures(ssGSEA_score_df = .x,
                                      tumor_type = .y)) %>% 
    cowplot::plot_grid(plotlist = ., ncol = 3)


```


### Single-cell NB-FOXR2

Define ssGSEA calculation & plotting functions:

<details>

```{r sc-gsea-functions}

run_gsea <- function(seurat, alias,
                       genelist, genelist_name){
    
    # Note: bin_by is merely used to split the ssGSEA calculation for computational tractability
    # Predictions are later merged together
    # Selection of variable does not matter to result 

    genelist_intersect <- map(genelist,
                              ~ intersect(rownames(seurat),.x))
    
    message(glue("The following genes were not found in seurat object {alias}"))
    map(genelist,
        ~ setdiff(.x, rownames(seurat))) %>% print()
    
    bin_ssgsea(seurat,
               bin_by = "seurat_clusters",
               summary_type = "single_cells",
               out_prefix = glue("{out}/{alias}_{genelist_name}"), 
               genelist = genelist,
               save_leading_edge = F)
    
}


get_scores <- function(seurat, 
                       alias,
                       metadata = NULL,
                       genelist_name) {
    
    scores <- data.table::fread(glue("{out}/{alias}_{genelist_name}.single_cells.ssgsea_scores.txt", 
                                     data.table = FALSE)) %>%
                column_to_rownames(var = "Signature") %>% 
                t() %>% 
                as.data.frame() %>% 
                rownames_to_column("Cell")
    
    if(!is.null(metadata)){
        scores <- scores %>% left_join(., metadata[[alias]])
    }
    
    return(scores)
    
}

plot_dist_score <- function(scores,
                            alias,
                            comparators,
                            palette) {

    scores %>%
        filter(inferCNV_call == "Tumor") %>%
        pivot_longer(cols = comparators,
                     names_to = "Signature",
                     values_to = "ssGSEA_score") %>%
        ggplot(aes(x = ssGSEA_score, fill = Signature)) +
        geom_density(alpha = 0.7) +
        ggtitle(alias) +
        scale_fill_manual(values = palette)
}

plot_dist_score_hg <- function(scores,
                            alias,
                            comparators,
                            palette,
                            mal_only = T,
                            mal_cell_ids = NULL) {
    
        if(mal_only) {
            if(is.null(mal_cell_ids)) {
                
                scores <- scores %>% 
                    filter(Malignant_normal == "Malignant")
                
            } else {
                 
                scores <- scores %>% 
                    filter(Cell %in% mal_cell_ids)
            }
        }
    
    scores %>% 
        pivot_longer(cols = comparators, 
                     names_to = "Signature", 
                     values_to = "ssGSEA_score") %>% 
        ggplot(aes(x = ssGSEA_score, fill = Signature)) +
        geom_density(alpha = 0.7) +
        ggtitle(alias) +
        scale_fill_manual(values = palette)
}

plot_dist_score_ridge <- function(scores,
                            alias,
                            comparators,
                            palette,
                            y_split_var,
                            color_var,
                            scale = 0.9) {

    scores %>%
        filter(Malignant_status == "Tumor") %>%
        pivot_longer(cols = comparators,
                     names_to = "Signature",
                     values_to = "ssGSEA_score") %>%
        ggplot(aes(x = ssGSEA_score,
                   y = .data[[y_split_var]],
                   fill = .data[[color_var]])) +
        geom_density_ridges(alpha = 0.7,
                            scale = scale) +
        ggtitle(alias) +
        scale_fill_manual(values = palette)
}


plot_dist_score_ridge_hg <- function(scores,
                            alias,
                            comparators,
                            palette,
                            y_split_var,
                            color_var, 
                            scale = 0.9,
                            mal_only = T,
                            mal_cell_ids = NULL) {
    
        if(mal_only) {
            if(is.null(mal_cell_ids)) {
                
                scores <- scores %>% 
                    filter(Malignant_normal == "Malignant")
                
            } else {
                 
                scores <- scores %>% 
                    filter(Cell %in% mal_cell_ids)
            }
        }

    
        scores %>% 
            pivot_longer(cols = comparators, 
                     names_to = "Signature", 
                     values_to = "ssGSEA_score") %>% 
            ggplot(aes(x = ssGSEA_score, 
                       y = .data[[y_split_var]], 
                       fill = .data[[color_var]])) +
            geom_density_ridges(alpha = 0.7,
                                scale = scale) +
            ggtitle(alias) +
            scale_fill_manual(values = palette)
}


```

</details>

Load single-cell tumor metadata and Seurat objects.

```{r get-meta-objects, message = F, warning = F, cache = F}

meta_sc <- read_tsv(here("output/00/metadata_patients_sc.tsv"))
sc_samples_nb_foxr2 <- meta_sc %>% filter(Group == "NB-FOXR2") %>% pull(ID)

# Function to load an RData file and return it
loadRData <- function(fileName){
  
  load(fileName)
  get(ls()[ls() != "fileName"])
  
}

# Load per-sample seurat objects
samples_sc <- map(sc_samples_nb_foxr2, ~ loadRData(meta_sc[meta_sc$ID == .x, ]$Path))
names(samples_sc) <- sc_samples_nb_foxr2

# Load merged object metadata
load(here("output/04/merged_sc_meta.Rda"))

# correct a discrepancy in how barcodes are named for the multiome sample
samples_sc$`P-6778_S-10155`@meta.data$cell.barcode <- samples_sc$`P-6778_S-10155`@meta.data$gex_barcode
merged_sc_meta <- merged_sc_meta %>%
    mutate(cell.barcode = ifelse(is.na(cell.barcode), gex_barcode, cell.barcode))

# Load samples
samples_sc <- map(samples_sc, function(seurat) {
    
    # sanity check that cell order is preserved in each metadata slot
    merged_barcodes <- merged_sc_meta %>% filter(cell.barcode %in% seurat@meta.data$cell.barcode) %>% pull(cell.barcode)
    all(seurat@meta.data$cell.barcode == merged_barcodes)
    
    return(seurat)
    
})

# Add in FOXR2 tumor metadata computed in previous R markdown
load(here("output/06.2/seurat_tumor_metadata.Rda"))
samples_sc <- map2(samples_sc, seurat_tumor_metadata,
                     function(seurat, meta) {
                         seurat@meta.data <- meta %>% 
                             column_to_rownames("Cell")
                         return(seurat)
                     })

```

Run ssGSEA scoring.

```{r run_ssgsea_hg, eval = FALSE}

# Run ssGSEA
purrr::imap(samples_sc,
          ~ run_gsea(seurat = .x,
                       alias = .y,
                       genelist = sigs_list_ssgsea_hg,
                       genelist_name = "tumor_sigs_comparator_consensus"))

```

```{r get_scores_hg, cache = F}

scores_hg <- purrr::imap(samples_sc, 
            ~ get_scores(seurat = .x,
                         alias = .y,
                         metadata = seurat_tumor_metadata,
                         genelist_name = "tumor_sigs_comparator_consensus"))
names(scores_hg) <- names(samples_sc)

```

Plot the distribution of scores in malignant cells.

```{r plot_ssgsea_enrichment_hg, fig.height=10, fig.width=30, cache = F}

purrr::imap(scores_hg,
            ~ plot_dist_score_hg(scores = .x,
                              alias = .y,
                              comparators = names(sigs_list_ssgsea_hg),
                              palette = palette_groups)) %>% 
    plot_grid(plotlist = .,
              align = "h",
              ncol = 3)

```


## Signature QC: heatmap

To summarize our validation efforts, here we generate a Z-scored heatmap of 
signature gene expression across bulk human pediatric brain tumors.

```{r load-hg-for-heatmap}

# Load genes

load(glue("{out}/sigs_list_ssgsea_mm.Rda"))

sig_genes <- sigs_list_ssgsea[["NB-FOXR2"]]
etmr_genes <- sigs_list_ssgsea[["ETMR"]]
dipg_genes <- sigs_list_ssgsea[["DIPG-H3K27M"]]
wnt_mb_genes <- sigs_list_ssgsea[["MB-WNT"]]

gene_split <- data.frame("Gene" = c(sig_genes, etmr_genes, wnt_mb_genes, dipg_genes),
                        "Group" = c(rep("NB-FOXR2", length(sig_genes)),
                                    rep("ETMR", length(etmr_genes)),
                                    rep("MB-WNT", length(wnt_mb_genes)),
                                    rep("DIPG", length(dipg_genes)))
                        )

# Convert gene symbols mouse to human using cached biomaRt reference
load(here("data/singlecell/references_genome/biomaRt_mm_to_hg_lds.Rda"))
gene_split_hg <- genes_lds %>% filter(MGI.symbol %in% gene_split$Gene) %>% 
    rename(Gene = MGI.symbol) %>% 
    rename(Gene_hg = HGNC.symbol) %>% 
    left_join(gene_split, .) %>% 
    select(-Gene)

meta <- read_tsv(here("output/00/metadata_patients_NGS.tsv"))
meta_bulk <- meta %>% filter(RNAseq == "Y")

korshunov <- meta_bulk %>% 
    filter(Group == "NB-FOXR2") %>% 
    filter(grepl(x = .$Source, pattern = "Korshunov")) %>% 
    .$ID %>%
    gsub(pattern = "$", replacement = ".0")

in_house <- meta_bulk %>% 
    filter(Group == "NB-FOXR2") %>% 
    filter(!grepl(x = .$Source, pattern = "Korshunov")) %>% 
    .$ID 

MBWNT_bulk_sample_names <- meta_bulk %>% 
    filter(Group == "MB-WNT") %>% 
    .$ID

ETMR_bulk_sample_names <- meta_bulk %>% 
    filter(Group == "ETMR") %>% 
    .$ID

DIPG_bulk_sample_names <- meta_bulk %>% 
    filter(Group == "DIPG-H3K27M") %>% 
    .$ID 

hg_nbfoxr2 <- extract_pipeline_counts(path = here("data/RNAseq/pipeline_l3/2023-05-test_pbt/counts/Ensembl.ensGene.exon.norm.tsv.gz"),
                                        goi = gene_split_hg$Gene_hg)  %>% 
    filter(sample %in% c(korshunov,
                         in_house,
                         MBWNT_bulk_sample_names,
                         ETMR_bulk_sample_names,
                         DIPG_bulk_sample_names))

hg_nbfoxr2 <- hg_nbfoxr2 %>% 
    mutate(Group = case_when(sample %in% korshunov ~ "NB-FOXR2 In-house",
                             sample %in% in_house ~ "NB-FOXR2 Korshunov",
                             sample %in% MBWNT_bulk_sample_names ~ "MB-WNT",
                             sample %in% ETMR_bulk_sample_names ~ "ETMR",
                             sample %in% DIPG_bulk_sample_names ~ "DIPG"))

# Subset gene_split_hg to the genes which were retrieved by extract_pipeline_counts
gene_split_hg <- gene_split_hg %>% 
    filter(Gene_hg %in% (hg_nbfoxr2$gene_symbol %>% unique))

```

Produce normalized counts matrix:

```{r norm-counts-hg, fig.height=13, fig.width=12}

# For genes with multiple ensembl IDs for a given symbol, select the symbol with
# highest gene_expression. If two have the same gene_expression, select randomly
mat <- hg_nbfoxr2 %>%
    group_by(gene_symbol, sample) %>% 
    slice_max(order_by = gene_expression, with_ties = F) %>% 
    select(gene_symbol, sample, gene_expression) %>%
    pivot_wider(names_from = sample, values_from = gene_expression) %>%
    tibble::column_to_rownames("gene_symbol") %>%
    as.matrix

sample_split <- hg_nbfoxr2 %>%
    select(sample, Group) %>%
    distinct()

```

Scale each row (gene) by Z score, group together NB-FOXR2 from in-house and 
Korshunov datasets, and annotate genes of interest on the right side. We rasterize the heatmap for Adobe Illustrator.

```{r heatmap-hg-scaled-row-zscore-groupnbfoxr2-annogenes, fig.height=13, fig.width=12}

mat_row_zscore <- apply(mat, 1,
                        scale) %>% t()
colnames(mat_row_zscore) <- colnames(mat)

sample_split_groupnbfoxr2 <- sample_split %>% 
    mutate(Group = case_when(Group == "NB-FOXR2 In-house" ~ "NB-FOXR2",
                             Group == "NB-FOXR2 Korshunov" ~ "NB-FOXR2",
                             TRUE ~ as.character(Group)))

m <- mat_row_zscore[gene_split_hg$Gene_hg,]

highlight <- c("NKX2-1", "FOXR2", "FOXG1", "LHX6", "ETS1", "DLX5", "NXPH1", "GPD1", "FAM163A", "DNM3", "FERMT1", 
               "LIN28A", "LIN28B", "HES5", "BEND3",
               "NKD1", "TBR1", "DKK2", "DKK4", "ZIC1", "ZIC3", "ZIC5",
               "MBP", "MOBP", "BCAS1", "MOG", "AQP4", "GFAP", "ALDH1L1")

ha <- rowAnnotation(an = anno_mark(at = which(rownames(m) %in% highlight),
                                   labels = rownames(m)[rownames(m) %in% highlight]))

Heatmap(m, use_raster = TRUE, raster_quality = 5,
        right_annotation = ha,
        cluster_rows = F, cluster_columns = F,
        col = circlize::colorRamp2(colors = c("royalblue3", "antiquewhite", "red2"),
                                  breaks = c(-3, 0, 3)),
        row_split = factor(gene_split_hg$Group, levels = c("NB-FOXR2", "ETMR", "MB-WNT", "DIPG")),
        column_split = factor(sample_split_groupnbfoxr2$Group, levels = c("NB-FOXR2",
                                                             "ETMR", "MB-WNT", "DIPG")),
        row_gap = unit(3, "mm"), column_gap = unit(3, "mm"),
        show_row_names = FALSE)

```

## ssGSEA in mouse models

Here, we finally use the generated and quality-checked consensus tumor-specific
signatures to score cells in single-cell profiles of our mouse models.

### Single-cell

Run ssGSEA with signatures.

```{r run_ssgsea, eval = FALSE}

# Run ssGSEA
purrr::imap(seurat_mouse,
           ~ run_gsea(seurat = .x,
                        alias = .y,
                        genelist = sigs_list_ssgsea,
                        genelist_name = "tumor_sigs_comparator_consensus"))

```

Retrieve scores saved in output.

```{r get_scores}

scores <- purrr::imap(seurat_mouse,
            ~ get_scores(seurat = .x,
                         alias = .y,
                         metadata = seurat_mouse_metadata,
                         genelist_name = "tumor_sigs_comparator_consensus"))
names(scores) <- names(seurat_mouse)

```

Ridge plot split by projected cell type of each single cell in mouse samples 
(Glia, Neurons).

```{r plot_ssgsea_enrichment_celltype, fig.height = 7, fig.width=30, cache = F}

# Fix palette names so it matches the group variables in scores object
names(palette_groups)[which(names(palette_groups) == "HGG-H3.3G34R/V")] <- "HGG-H3.3G34R.V"

purrr::imap(scores,
            ~ plot_dist_score_ridge(scores = .x %>% filter(broad_label %in% c("Neuron", "Glia")),
                              alias = .y,
                              comparators = names(sigs_list_ssgsea),
                              palette = palette_groups,
                              y_split_var = "broad_label",
                              color_var = "Signature",
                              scale = 0.6))  %>%
    plot_grid(plotlist = .,
              align = "h",
              ncol = 3)

```

In the publication, we separate the plot rows for neurons (main figures) and glia 
(supplementary figures) in Adobe Illustrator.

### Joint UMAP of ssGSEA score in malignant neurons

Load joint object of all mouse model samples and metadata produced in previous R markdown `07.1`.

```{r load-joint}

load(here("data/singlecell/integrations/mouse_naive_join/output/seurat_joint.Rda"))
seurat_joint$Joint_cluster <- Idents(seurat_joint)

# For now, we will not use the ATAC, so let's make a "light" version of the object
seurat_joint[['ATAC']] <- NULL
seurat_joint[['peaks']] <- NULL
seurat_joint[['promoters']] <- NULL

# Sanity check: # cells in joint sample = total across individual samples
map(seurat_mouse, ncol) %>% unname() %>% as.numeric() %>% sum() == ncol(seurat_joint)

load(here("output/07.1/seurat_mouse_metadata_joint.Rda"))
seurat_joint@meta.data <- seurat_mouse_metadata_joint %>%
                             column_to_rownames("Cell")


```

Add ssGSEA scores per cell to joint object metadata and save the metadata dataframe.

```{r add-scores-meta-joint}

joint_scores <- imap(scores,
    function(df, sample_name) {
        df %>% 
            dplyr::rename("ssGSEA_NBFOXR2" = "NB-FOXR2") %>% 
            dplyr::rename("ssGSEA_ETMR" = "ETMR") %>%
            dplyr::rename("ssGSEA_DIPG" = "DIPG-H3K27M") %>%
            dplyr::rename("ssGSEA_MB_WNT" = "MB-WNT") %>%
            mutate("Sample" = sample_name) %>% 
            select("gex_barcode", "ssGSEA_NBFOXR2", "ssGSEA_ETMR", "ssGSEA_DIPG", "ssGSEA_MB_WNT", "Sample", "orig.ident")
    }) %>% Reduce(rbind, .) %>% 
    left_join(by = c("gex_barcode", "Sample"), seurat_joint@meta.data) %>% 
    select(ssGSEA_NBFOXR2, ssGSEA_ETMR, ssGSEA_DIPG, ssGSEA_MB_WNT) 

rownames(joint_scores) <- rownames(seurat_joint@meta.data)

seurat_joint <- AddMetaData(seurat_joint, joint_scores)

save(joint_scores, file = glue("{out}/joint_mouse_ssGSEA_scores_meta.Rda"))

```

Remove No consensus cells and doublets from the joint object for plotting purpose.

```{r save-joint-obj-plot}

seurat_joint <- subset(seurat_joint, subset = broad_label != "No consensus" & !is.na(broad_label))

```

In order to set same score color legend for all of the UMAPs, we check the range of values for each signature score.

```{r set-range-legend}

range(seurat_joint$ssGSEA_NBFOXR2)
range(seurat_joint$ssGSEA_MB_WNT)
range(seurat_joint$ssGSEA_ETMR)
range(seurat_joint$ssGSEA_DIPG)

```

* Max value: max(ssGSEA_DIPG)
* Min value: min(ssGSEA_MB_WNT)

We set legend to be between these two values for all ssGSEA UMAP plots.

For the publication figure, we will superimpose the neurons colored by 
ssGSEA score over a grey UMAP showing all cells. We will produce each UMAP
separately in this R markdown and superimpose in Adobe Illustrator.

Produce grey UMAP with malignant and normal cells together:

```{r joint-umap-grey}

p_grey <- DimPlot(seurat_joint, pt.size = 1,
                  group.by = "is_cell",
                  cols = "grey80") + square_theme + large_text

ggrastr::rasterise(p_grey, dpi = 2000)

```

Extract UMAP plot axis ranges (to create the next plot in exact alignment).

```{r extract-joint-umap-plot-axis-ranges}

# x-range
x_range <- layer_scales(p_grey)$x$range$range

# y-range
y_range <- layer_scales(p_grey)$y$range$range

```

Save joint object subset to malignant neurons only:

```{r save-joint-obj-plot-mal-neur-only}

seurat_joint_mal_neurons <- subset(seurat_joint, subset = broad_label == "Neuron")

```

Plot ssGSEA score in malignant neurons only, ordering points by score. Rasterize these plots for Adobe Illustrator.

```{r umap-ssgsea-mouse-joint-mal-neur-only, fig.height=6, fig.width=7}

## FOXR2 ssgsea
p1 <- FeaturePlot(seurat_joint_mal_neurons, pt.size = 1, features = "ssGSEA_NBFOXR2", order = T) +
  scale_color_gradientn(colors = c("darkblue", "white", "red"),
                         limits = c(min(seurat_joint$ssGSEA_MB_WNT), max(seurat_joint$ssGSEA_DIPG))) + 
    xlim(x_range) + ylim(y_range) +
    square_theme + large_text

ggrastr::rasterise(p1, dpi = 2000)

## MB-WNT ssgsea
p2 <- FeaturePlot(seurat_joint_mal_neurons, pt.size = 1, features = c("ssGSEA_MB_WNT"), order = T) +
  scale_color_gradientn(colors = c("darkblue", "white", "red"),
                         limits = c(min(seurat_joint$ssGSEA_MB_WNT), max(seurat_joint$ssGSEA_DIPG))) + 
    xlim(x_range) + ylim(y_range) +
    square_theme + large_text

ggrastr::rasterise(p2, dpi = 2000)

## ETMR ssgsea
p3 <- FeaturePlot(seurat_joint_mal_neurons, pt.size = 1, features = c("ssGSEA_ETMR"), order = T) +
  scale_color_gradientn(colors = c("darkblue", "white", "red"),
                         limits = c(min(seurat_joint$ssGSEA_MB_WNT), max(seurat_joint$ssGSEA_DIPG))) + 
    xlim(x_range) + ylim(y_range) +
    square_theme + large_text

ggrastr::rasterise(p3, dpi = 2000)

## DIPG ssgsea
p4 <- FeaturePlot(seurat_joint_mal_neurons, pt.size = 1, features = c("ssGSEA_DIPG"), order = T) +
  scale_color_gradientn(colors = c("darkblue", "white", "red"),
                         limits = c(min(seurat_joint$ssGSEA_MB_WNT), max(seurat_joint$ssGSEA_DIPG))) + 
    xlim(x_range) + ylim(y_range) +
    square_theme + large_text

ggrastr::rasterise(p4, dpi = 2000)

```


<!-- END MATTER, insert reproducibility info -->

```{r footer, echo = FALSE, results = 'asis', warning = FALSE, cache = FALSE}

# Knit child document with header
res <- knitr::knit_child(here("include", "footer.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF END MATTER -->