---
title: "08 - Extracranial neuroblastoma"
author: "Selin Jessa [[selin.jessa@mail.mcgill.ca](mailto:selin.jessa@mail.mcgill.ca)] and Bhavyaa Chandarana [[bhavyaa.chandarana@mail.mcgill.ca](mailto:bhavyaa.chandarana@mail.mcgill.ca)]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: yes
    code_folding: show
    theme: flatly
    css: ../include/style.css
    toc: yes
    toc_depth: 4
    number_sections: true
    df_print: paged
    includes:
      before_body: ../include/header.html
      after_body:  ../include/footer.html
---

<!-- FRONT MATTER, insert configuration info -->

```{r header, echo = FALSE, results = 'asis', warning = FALSE}

# Index of the document
# ...determines name of the subfolder of `outputs` and `figures`
doc_id <- "08"

suppressMessages(library(here))

# Knit child document with header
res <- knitr::knit_child(here("include", "header.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF FRONT MATTER -->


# Overview

This document compares NB-FOXR2 with extracranial neuroblastoma samples
stratified by stage and FOXR2 expression.

# Libraries

```{r libs, warning = FALSE, message = FALSE, echo = TRUE, cache = FALSE}

library(here)
library(magrittr)
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(stringr)
library(glue)
library(purrr)
library(ggplot2)
library(ggrepel)
library(tibble)
library(cowplot)
library(Seurat)
library(data.table)
library(MetBrewer)
library(dendextend)
library(readxl)
library(GSVA)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
library(ComplexHeatmap)

source(here("include/style.R"))
source(here("code/functions/RNAseq.R"))
source(here("code/functions/scRNAseq.R"))
source(here("code/functions/ssGSEA.R"))

ggplot2::theme_set(theme_min())

square_theme <- theme(aspect.ratio = 1)
large_text <- theme(text = element_text(size = 15))

conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::first)
conflicted::conflicts_prefer(dplyr::last)
conflicted::conflicts_prefer(dplyr::combine)
conflicted::conflicts_prefer(dplyr::intersect)
conflicted::conflicts_prefer(base::setdiff)
conflicted::conflicts_prefer(clusterProfiler::slice)
conflicted::conflicts_prefer(dplyr::rename)
conflicted::conflicts_prefer(dplyr::desc)
conflicted::conflicts_prefer(dplyr::filter)

```

# Load EC-NB samples

## Gartlgruber et al. 2021 (n=579)

Publication: [Gartlgruber et al. *Nature Cancer* 2021](https://pubmed.ncbi.nlm.nih.gov/35121888/)

### Load data

This data was downloaded from a 
[Shiny app]((https://nbseb087.dkfz.de/project_NB_SE_viz/#shiny-tab-faq)) 
provided by the authors of the study.

Data was provided as un-normalized bulk counts and per-tumor metadata.

Load DESeq normalized counts produced in in R markdown document `03` for plotting
gene expression.

```{r load-norm-counts}

# Load files and remove unused ones to save space
load(here("output/03/Gartlgruber_et_al_counts.Rda"))
rm(ecnb_counts_norm)
rm(ecnb_counts_vst)

```

### Add FOXR2 positivity to metadata

Here, we label each tumor as FOXR2 positive or negative in per-tumor metadata 
based on normalized expression of FOXR2. 
We consider tumors with FOXR2 DESeq normalized expression > 2 to be FOXR2 positive. 

```{r add-foxr2-positivity-meta}

# Set threshold of FOXR2+ expression used throughout this document
threshold <- 2

ecnb_foxr2_counts <- ecnb_counts_tidy %>%
    filter(gene_symbol == "FOXR2")

foxr2_pos_samples <- ecnb_foxr2_counts %>% 
    filter(gene_expression > threshold) %>% 
    .$sample

ecnb_foxr2_counts <- ecnb_foxr2_counts %>% mutate(FOXR2_positive = case_when(sample %in% foxr2_pos_samples ~ "Y",
                                      T ~ "N"))

# Print the number of FOXR2 positive and negative samples
ecnb_foxr2_counts %>% count(FOXR2_positive, sort = F)

# Print the number of FOXR2 positive and negative samples in each tumor stage
ecnb_foxr2_counts %>% count(Stage, FOXR2_positive, sort = F)

ecnb_counts_tidy <- ecnb_counts_tidy %>% 
    mutate(FOXR2_positive = case_when(sample %in% foxr2_pos_samples ~ "FOXR2+",
                                      T ~ "FOXR2-"))

```

## TARGET cohort (n=128)

Downloaded and processed by Steven Hébert in April 2024. 

They were downloaded as count matrices from the public Genomic
686 Data Commons (GDC) cancer portal (https://portal.gdc.cancer.gov/) 
using the Cohort Builder, with options: 

* `project = TARGET-346 NBL`
* `Experimental Strategy = RNA-seq`
* `Access = open`

### Load data

Patients were subsetted to Risk = HR (high risk) and Stage = 3 or 4 (not 4S).
This resulted in a total of 128 patient samples.

Load patient sample metadata.

```{r load-ecnb-target-meta}

# Load, replace the empty value '-- with NA values and remove all-NA columns, 
# and rename id column to Sample
target_meta <- data.table::fread(here("data/RNAseq/external_data/TARGET_ECNB/clinical.tsv")) %>%
    apply(2, 
              function(col) { 
                  gsub(col, pattern = "'--", replacement = as.character(NA)) 
              }
          ) %>% 
    as.data.frame() %>% 
    select_if(~ !all(is.na(.))) %>% 
    dplyr::rename(Sample = case_submitter_id)

```

Load DESeq-normalized counts (for plotting gene expression).

```{r load-ecnb-target-norm-counts}

# load normalized counts
target_norm <- here("data/RNAseq/external_data/TARGET_ECNB/counts_ensembl_common/GENCODEv36.norm.tsv.gz") %>%
    read.table(header = T, row.names = 1, check.names = F, sep = "\t")

# Display first 5 row/col to check
target_norm[1:5, 1:5]

```

### Add FOXR2 positivity to metadata

Based on threshold = 2.

```{r add-foxr2-positivity-meta-target}

# DESeq norm expression threshold used for FOXR2 +/-
threshold <- 2

target_counts_tidy <- target_norm %>% 
    as.data.frame() %>% 
    rownames_to_column("Gene") %>% 
    separate(col = "Gene", sep = ":", into = c("ENS", "gene_symbol")) %>% 
    pivot_longer(cols = -c("ENS", "gene_symbol"), names_to = "ID", values_to = "gene_expression")
    
foxr2_pos_samples_target <- target_counts_tidy %>% 
    filter(gene_symbol == "FOXR2") %>% 
    filter(gene_expression > threshold) %>% 
    .$ID 

target_meta_foxr2 <- target_counts_tidy %>% 
    mutate(FOXR2_positive = case_when(ID %in% foxr2_pos_samples_target ~ "Y",
                                      T ~ "N")) %>% 
    select(ID, FOXR2_positive) %>% distinct

```

# FOXR2 coexpression with MGE TFs in EC-NB

## Gartlgruber et al. 2021

Stage 4, high-risk tumors only. (Samples with NA in either Risk or Stage metadata columns are removed.)

```{r FOXR2-coexp-ECNB, fig.height=4, fig.width=24}

tum_interest <- ecnb_counts_tidy %>% 
    filter(Risk == "HR") %>%
    filter(Stage == "4") %>% 
    filter(gene_symbol %in% c("FOXR2", "LHX6", "DLX5", "DLX6"))

# DLX5
foxr2_dlx5 <- tum_interest %>% 
    filter(gene_symbol %in% c("FOXR2", "DLX5")) %>% 
    pivot_wider(names_from = "gene_symbol", 
                values_from = "gene_expression", 
                id_cols = c("sample", "FOXR2_positive")) 

(cor_foxr2_dlx5 <- cor(foxr2_dlx5$FOXR2, foxr2_dlx5$DLX5, method = "pearson"))

(sig_dlx5 <- wilcox.test(foxr2_dlx5 %>% filter(FOXR2_positive == "FOXR2+") %>% .$DLX5,
                         foxr2_dlx5 %>% filter(FOXR2_positive == "FOXR2-") %>% .$DLX5))

dlx5_scatter <- foxr2_dlx5 %>% 
    ggplot(aes(x = FOXR2, y = DLX5, alpha = 0.2)) + 
    geom_point(alpha = 0.5, size = 3) + 
    ggtitle("DLX5 - EC-NB stage 4") +
    annotate("text", label = glue("r == {round(cor_foxr2_dlx5,4)}"),
             x = 470, y = 4000, parse = T)

dlx5_vln <- foxr2_dlx5 %>% 
    ggplot(aes(x = FOXR2_positive, y = DLX5, fill = FOXR2_positive)) + 
    geom_violin(scale = "width") + 
    ggtitle("DLX5 - EC-NB stage 4") + 
    no_legend() + scale_y_log10() + 
    stat_summary(fun.y=median, geom="crossbar", size=1, 
                 color="black", aes(width = 0.3)) +
    annotate("text", label = glue("p == {round(sig_dlx5$p.value,4)}"),
             x = 2, y = 4000, parse = T)


# DLX6
foxr2_dlx6 <- tum_interest %>% 
    filter(gene_symbol %in% c("FOXR2", "DLX6")) %>% 
    pivot_wider(names_from = "gene_symbol", 
                values_from = "gene_expression", 
                id_cols = c("sample", "FOXR2_positive")) 

(cor_foxr2_dlx6 <- cor(foxr2_dlx6$FOXR2, foxr2_dlx6$DLX6, method = "pearson"))

(sig_dlx6 <- wilcox.test(foxr2_dlx6 %>% filter(FOXR2_positive == "FOXR2+") %>% .$DLX6,
                         foxr2_dlx6 %>% filter(FOXR2_positive == "FOXR2-") %>% .$DLX6))

dlx6_scatter <- foxr2_dlx6 %>% 
    ggplot(aes(x = FOXR2, y = DLX6, alpha = 0.2)) + 
    geom_point(alpha = 0.5, size = 3) + ggtitle("DLX6 - EC-NB stage 4") +
    annotate("text", label = glue("r == {round(cor_foxr2_dlx6,4)}"),
             x = 470, y = 3000, parse = T)

dlx6_vln <- foxr2_dlx6 %>% 
    ggplot(aes(x = FOXR2_positive, y = DLX6, fill = FOXR2_positive)) + 
    geom_violin(scale = "width") + 
    ggtitle("DLX6 - EC-NB stage 4") + 
    no_legend() + scale_y_log10() + 
    stat_summary(fun.y=median, geom="crossbar", size=1, 
                 color="black", aes(width = 0.3))  +
    annotate("text", label = glue("p == {round(sig_dlx6$p.value,4)}"),
             x = 2, y = 4000, parse = T)



# LHX6
foxr2_lhx6 <- tum_interest %>% 
    filter(gene_symbol %in% c("FOXR2", "LHX6")) %>% 
    pivot_wider(names_from = "gene_symbol", 
                values_from = "gene_expression", 
                id_cols = c("sample", "FOXR2_positive")) 

(cor_foxr2_lhx6 <- cor(foxr2_lhx6$FOXR2, foxr2_lhx6$LHX6, method = "pearson"))

(sig_lhx6 <- wilcox.test(foxr2_lhx6 %>% filter(FOXR2_positive == "FOXR2+") %>% .$LHX6,
                         foxr2_lhx6 %>% filter(FOXR2_positive == "FOXR2-") %>% .$LHX6))

lhx6_scatter <- foxr2_lhx6 %>% 
    ggplot(aes(x = FOXR2, y = LHX6, alpha = 0.2)) + 
    geom_point(alpha = 0.5, size = 3) + 
    ggtitle("LHX6 - EC-NB stage 4") +
    annotate("text", label = glue("r == {round(cor_foxr2_lhx6,4)}"),
             x = 470, y = 600, parse = T)

lhx6_vln <- foxr2_lhx6 %>% 
    ggplot(aes(x = FOXR2_positive, y = LHX6, fill = FOXR2_positive)) + 
    geom_violin(scale = "width") + 
    ggtitle("LHX6 - EC-NB stage 4") + 
    no_legend() + 
    stat_summary(fun.y=median, geom="crossbar", size=1, 
                 color="black", aes(width = 0.3))  +
    annotate("text", label = glue("p == {round(sig_lhx6$p.value,4)}"),
             x = 2, y = 600, parse = T)


cowplot::plot_grid(dlx5_scatter, dlx5_vln,
                                dlx6_scatter, dlx6_vln,
                                lhx6_scatter, lhx6_vln,
                   nrow = 1, align = "hv")

```

## TARGET cohort

FOXR2 + threshold: DESEq normalized value "2".

```{r FOXR2-coexp-TARGET, fig.height=4, fig.width=24}

tum_interest <- target_counts_tidy %>% 
    filter(gene_symbol %in% c("FOXR2", "LHX6", "DLX5", "DLX6")) %>%
    left_join(., target_meta_foxr2, by = "ID") %>% 
    dplyr::rename("sample" = "ID") %>% 
    mutate(FOXR2_positive = case_when(FOXR2_positive == "Y" ~ "FOXR2+",
                                      FOXR2_positive == "N" ~ "FOXR2-",
                                      TRUE ~ NA))

# DLX5
foxr2_dlx5 <- tum_interest %>% 
    filter(gene_symbol %in% c("FOXR2", "DLX5")) %>% 
    pivot_wider(names_from = "gene_symbol", 
                values_from = "gene_expression", 
                id_cols = c("sample", "FOXR2_positive")) 

(cor_foxr2_dlx5 <- cor(foxr2_dlx5$FOXR2, foxr2_dlx5$DLX5, method = "pearson"))

(sig_dlx5 <- wilcox.test(foxr2_dlx5 %>% filter(FOXR2_positive == "FOXR2+") %>% .$DLX5,
                         foxr2_dlx5 %>% filter(FOXR2_positive == "FOXR2-") %>% .$DLX5))

dlx5_scatter <- foxr2_dlx5 %>% 
    ggplot(aes(x = FOXR2, y = DLX5, alpha = 0.2)) + 
    geom_point(alpha = 0.5, size = 3) + 
    ggtitle("DLX5 - EC-NB stage 4") +
    annotate("text", label = glue("r == {round(cor_foxr2_dlx5,4)}"),
            x = 900, y = 5500, parse = T)

dlx5_vln <- foxr2_dlx5 %>% 
    ggplot(aes(x = FOXR2_positive, y = DLX5, fill = FOXR2_positive)) + 
    geom_violin(scale = "width") + 
    ggtitle("DLX5 - EC-NB stage 4") + 
    no_legend() + scale_y_log10() + 
    stat_summary(fun.y=median, geom="crossbar", size=1, 
                 color="black", aes(width = 0.3)) +
    annotate("text", label = glue("p == {round(sig_dlx5$p.value,4)}"),
             x = 2, y = 5500, parse = T)


# DLX6
foxr2_dlx6 <- tum_interest %>% 
    filter(gene_symbol %in% c("FOXR2", "DLX6")) %>% 
    pivot_wider(names_from = "gene_symbol", 
                values_from = "gene_expression", 
                id_cols = c("sample", "FOXR2_positive")) 

(cor_foxr2_dlx6 <- cor(foxr2_dlx6$FOXR2, foxr2_dlx6$DLX6, method = "pearson"))

(sig_dlx6 <- wilcox.test(foxr2_dlx6 %>% filter(FOXR2_positive == "FOXR2+") %>% .$DLX6,
                         foxr2_dlx6 %>% filter(FOXR2_positive == "FOXR2-") %>% .$DLX6))

dlx6_scatter <- foxr2_dlx6 %>% 
    ggplot(aes(x = FOXR2, y = DLX6, alpha = 0.2)) + 
    geom_point(alpha = 0.5, size = 3) + ggtitle("DLX6 - EC-NB stage 4") +
    annotate("text", label = glue("r == {round(cor_foxr2_dlx6,4)}"),
              x = 900, y = 3500, parse = T)

dlx6_vln <- foxr2_dlx6 %>% 
    ggplot(aes(x = FOXR2_positive, y = DLX6, fill = FOXR2_positive)) + 
    geom_violin(scale = "width") + 
    ggtitle("DLX6 - EC-NB stage 4") + 
    no_legend() + scale_y_log10() + 
    stat_summary(fun.y=median, geom="crossbar", size=1, 
                 color="black", aes(width = 0.3))  +
    annotate("text", label = glue("p == {round(sig_dlx6$p.value,4)}"),
              x = 2, y = 3500, parse = T)



# LHX6
foxr2_lhx6 <- tum_interest %>% 
    filter(gene_symbol %in% c("FOXR2", "LHX6")) %>% 
    pivot_wider(names_from = "gene_symbol", 
                values_from = "gene_expression", 
                id_cols = c("sample", "FOXR2_positive")) 

(cor_foxr2_lhx6 <- cor(foxr2_lhx6$FOXR2, foxr2_lhx6$LHX6, method = "pearson"))

(sig_lhx6 <- wilcox.test(foxr2_lhx6 %>% filter(FOXR2_positive == "FOXR2+") %>% .$LHX6,
                         foxr2_lhx6 %>% filter(FOXR2_positive == "FOXR2-") %>% .$LHX6))

lhx6_scatter <- foxr2_lhx6 %>% 
    ggplot(aes(x = FOXR2, y = LHX6, alpha = 0.2)) + 
    geom_point(alpha = 0.5, size = 3) + 
    ggtitle("LHX6 - EC-NB stage 4") +
    annotate("text", label = glue("r == {round(cor_foxr2_lhx6,4)}"),
             x = 900, y = 900, parse = T)

lhx6_vln <- foxr2_lhx6 %>% 
    ggplot(aes(x = FOXR2_positive, y = LHX6, fill = FOXR2_positive)) + 
    geom_violin(scale = "width") + 
    ggtitle("LHX6 - EC-NB stage 4") + 
    no_legend() + 
    stat_summary(fun.y=median, geom="crossbar", size=1, 
                 color="black", aes(width = 0.3))  +
    annotate("text", label = glue("p == {round(sig_lhx6$p.value,4)}"),
             x = 2, y = 900, parse = T)


cowplot::plot_grid(dlx5_scatter, dlx5_vln,
                    dlx6_scatter, dlx6_vln,
                    lhx6_scatter, lhx6_vln,
                   nrow = 1, align = "hv")


```


<!-- END MATTER, insert reproducibility info -->

```{r footer, echo = FALSE, results = 'asis', warning = FALSE, cache = FALSE}

# Knit child document with header
res <- knitr::knit_child(here("include", "footer.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF END MATTER -->
