---
title: "09 - FOXR2 CUT&RUN in FOXR2 p53LOF mouse models"
author: "Bhavyaa Chandarana [[bhavyaa.chandarana@mail.mcgill.ca](mailto:bhavyaa.chandarana@mail.mcgill.ca)] and Steven Hébert [[steven.hebert@ladydavis.ca](mailto:steven.hebert@ladydavis.ca)]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    self_contained: yes
    keep_md: yes
    code_folding: show
    theme: flatly
    css: ../include/style.css
    toc: yes
    toc_depth: 4
    number_sections: true
    df_print: paged
    includes:
      before_body: ../include/header.html
      after_body:  ../include/footer.html
---

<!-- FRONT MATTER, insert configuration info -->

```{r header, echo = FALSE, results = 'asis', warning = FALSE}

# Index of the document
# ...determines name of the subfolder of `outputs` and `figures`
doc_id <- "09"

suppressMessages(library(here))

# Knit child document with header
res <- knitr::knit_child(here("include", "header.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF FRONT MATTER -->


# Overview

In this document, we analyze FOXR2 CUT&RUN (similar to ChIP) performed in n=3 
cell lines derived from Foxr2 p53 LOF murine models.

Note that throughout the code we refer to the CUT&RUN data as "ChIP" due
to its similarity in the analysis stage.

# Libraries

```{r libs, warning = FALSE, message = FALSE, echo = TRUE, cache = FALSE}

library(here)
library(magrittr)
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(stringr)
library(glue)
library(purrr)
library(ggplot2)
library(cowplot)
library(ComplexHeatmap)
library(GenomeInfoDb)
library(GenomicRanges)
library(GenomicFeatures)
library(rtracklayer)
library(Seurat)
library(Signac)
library(gprofiler2)
library(VennDiagram)

source(here("include/style.R"))
source(here("code/functions/scRNAseq.R"))
source(here("code/functions/RNAseq.R"))

ggplot2::theme_set(theme_min())


```

# Load data

FOXR2-V5 CUT&RUN (similar to ChIP) was run with two protocols "native" and "X-link",
with n=2 replicates each. 

A FOXR2-V5 experiment was also run in a cell line lacking V5-tagged insertion (n=2 
replicates). We also load it here to use as a control.

```{r load-peaks, cache=F}

chip_path <- here("data/ChIPseq/pipeline_genpipes/peaks")

# X-link FOXR2 peaks
rep1_FOXR2_peaks <- rtracklayer::import(glue("{chip_path}/AN24377_FOXR2_P53LOF_1_X-link_JAB1979A5-1.V5_peaks.narrowPeak"), format = "narrowPeak") 
rep2_FOXR2_peaks <- rtracklayer::import(glue("{chip_path}/AN24377_FOXR2_P53LOF_2_X-link_JAB1979A5-2.V5_peaks.narrowPeak"), format = "narrowPeak")

# X-link V5 peaks
rep1_V5_peaks <- rtracklayer::import(glue("{chip_path}/AH25351_G34R_1_X-link_JAB1979A5-3.V5_peaks.narrowPeak"), format = "narrowPeak")
rep2_V5_peaks <- rtracklayer::import(glue("{chip_path}/AH25351_G34R_2_X-link_JAB1979A5-4.V5_peaks.narrowPeak"), format = "narrowPeak")

# Create lists of peaks
chip_foxr2_peak_list <- list(rep1_FOXR2 = rep1_FOXR2_peaks,
               rep2_FOXR2 = rep2_FOXR2_peaks)

chip_V5_peak_list <- list(rep1_V5 = rep1_V5_peaks,
               rep2_V5 = rep2_V5_peaks)

# Remove non-standard chromosomes/contigs from each sample
standard_chrom_mm <- function(gr) { 
    gr <- gr[GenomicRanges::seqnames(gr) %in% c(1:19, 'X', 'Y')]
    GenomeInfoDb::seqlevels(gr) <- c(1:19, 'X', 'Y')
    return(gr)
}

chip_foxr2_peak_list <- map(chip_foxr2_peak_list, standard_chrom_mm)
chip_V5_peak_list <- map(chip_V5_peak_list, standard_chrom_mm)

```

# Define peaks

## Filter V5 peaks

We have control data from FOXR2-V5 CUT&RUN in a cell line lacking V5 insertion.
The peaks in this sample arise from binding regions of V5 antibody
in the genome. We use this as a background to filter out false positive peaks.

```{r compare-peak-count-v5, cache=F}

# Concatenate the V5 peaks 
all_V5_peaks <- c(chip_V5_peak_list[[1]], chip_V5_peak_list[[2]]) %>%
  sortSeqlevels %>%
  sort %>%
  reduce(min.gapwidth = 0)

# Remove V5 peaks from the FOXR2 peaks
chip_foxr2_peak_list_no_V5 <- chip_foxr2_peak_list
chip_foxr2_peak_list_no_V5[[1]] <- subsetByOverlaps(chip_foxr2_peak_list_no_V5[[1]], all_V5_peaks, invert = T, minoverlap = 10)
chip_foxr2_peak_list_no_V5[[2]] <- subsetByOverlaps(chip_foxr2_peak_list_no_V5[[2]], all_V5_peaks, invert = T, minoverlap = 10)

# Check the length of the peak lists (as a QC)
map(chip_foxr2_peak_list, length)
map(chip_foxr2_peak_list_no_V5, length)

# Create a narrowpeak file for the ChIP-seq FOXR2 samples after removing the V5 peaks 
write.table(chip_foxr2_peak_list_no_V5[[1]], file = glue("{out}/rep1_FOXR2_no_V5.narrowpeak"), sep = "\t", quote = F, row.names = F, col.names = T)
write.table(chip_foxr2_peak_list_no_V5[[2]], file = glue("{out}/rep2_FOXR2_no_V5.narrowpeak"), sep = "\t", quote = F, row.names = F, col.names = T)

```

## Intersect FOXR2 CUT&RUN & ATAC-seq peaks

We expect high-quality FOXR2 CUT&RUN peaks to overlap with open chromatin
(ATAC-seq peaks) derived from scMultiome of the same mouse model.

Here, we compute the overlap between these peaks, and use ATAC peaks
to filter CUT&RUN peaks. In the end, we retain the intersect of peaks between
ATAC-seq, and the two replicates of X-link FOXR2 CUT&RUN.

Define a function to plot triple venn diagram:

```{r triple-venn}

# Function to plot triple venn diagram from a list containing 3 GRanges objects
venn_from_gr_list <- function(gr_list, plot_title){
    
  overlap_all <- Reduce(function(x, y) subsetByOverlaps(x, y, minoverlap = 10), gr_list) %>% length()
  
  overlap_12 <- findOverlaps(gr_list[[1]], gr_list[[2]], minoverlap = 10)
  overlap_23 <- findOverlaps(gr_list[[2]], gr_list[[3]], minoverlap = 10)
  overlap_13 <- findOverlaps(gr_list[[1]], gr_list[[3]], minoverlap = 10)
  
  n_overlap_12 <- min(length(unique(queryHits(overlap_12))), length(unique(subjectHits(overlap_12))))
  n_overlap_23 <- min(length(unique(queryHits(overlap_23))), length(unique(subjectHits(overlap_23))))
  n_overlap_13 <- min(length(unique(queryHits(overlap_13))), length(unique(subjectHits(overlap_13))))
  
  venn.plot = draw.triple.venn(length(gr_list[[1]]),
                 length(gr_list[[2]]),
                 length(gr_list[[3]]),
                 n_overlap_12,
                 n_overlap_23,
                 n_overlap_13,
                 overlap_all,
                 category = c("rep1_FOXR2", "rep2_FOXR2", "ATAC"),
                 euler.d = T,
                 scaled = T,
                 fill = c("#33CC00", "#33CC00", "#009999"))

  grid.draw(venn.plot)
}
```

Load ATAC peaks:

```{r compare-chip-atac, cache=F}

# Load ATAC peaks produced from multiome processing pipeline
Foxr2_p53_r1_ATAC_peaks <- rtracklayer::import(here("data/singlecell/pipeline_scMultiome_mm/AN24377/ATAC/macs2.peaks.standard.blacklistremoved.narrowpeak"), format = "narrowPeak")

# Note that the blacklisted peaks were removed from list of peaks
# The list of blacklisted peaks comes from this set (from Signac package)
write.table(Signac::blacklist_mm10, file = glue("{out}/blacklist_mm10.tsv"), 
            sep = "\t", quote = F, row.names = F, col.names = F)

# Modify the blacklist_mm10 file to get a sorted bed files with names
system(glue("sort -Vk1,1 -k2,2 -k3,3 {out}/blacklist_mm10.tsv | 
               cut -f1-3,6 |
               sed 's/ /_/g' > {out}/blacklist_mm10.bed"))

# Remove non-standard chromosomes
Foxr2_p53_r1_ATAC_peaks <- standard_chrom_mm(Foxr2_p53_r1_ATAC_peaks) 

```

Plot number of peaks overlapping between CUT&RUN replicates 
and ATAC without any filtering:

```{r venn-overlap}

peak_list <- c(chip_foxr2_peak_list_no_V5, FOXR2_ATAC = Foxr2_p53_r1_ATAC_peaks)
venn_from_gr_list(peak_list, "venn_ChIP_ATAC_nofilter")

```

Filter CUT&RUN and ATAC peaks by quality metrics:

```{r filter-chip-atac}

# Add width column to the peaks
chip_foxr2_peak_list_no_V5 <- map(chip_foxr2_peak_list_no_V5, ~ { 
  .x$width <- width(.x)
  return(.x)
})

Foxr2_p53_r1_ATAC_peaks$width <- width(Foxr2_p53_r1_ATAC_peaks)

# Filter peaks of ChIP-seq and ATAC-seq
# Peaks from ChIP-seq and ATAC-seq have different value distribution 
# (signalValue, qValue), so different thresholds need to be used.
filtered_2_chip_foxr2_peak_list_no_V5 <- map(chip_foxr2_peak_list_no_V5,
        ~ .x[which(.x$qValue > 6 & .x$signalValue > 6 & .x$width < 10000),])
filtered_2_Foxr2_p53_r1_ATAC_peaks <- Foxr2_p53_r1_ATAC_peaks[Foxr2_p53_r1_ATAC_peaks$qValue > 4 &
                                                            Foxr2_p53_r1_ATAC_peaks$signalValue > 2 &
                                                            Foxr2_p53_r1_ATAC_peaks$width < 10000,]

```

Plot number of peaks overlapping between CUT&RUN replicates 
and ATAC after filtering.

This plot was adapted in Adobe Illustrator for the publication.

```{r venn-overlap-filtered}

### Check number of peaks overlapping
filtered_2_peak_list <- c(filtered_2_chip_foxr2_peak_list_no_V5, FOXR2_ATAC = filtered_2_Foxr2_p53_r1_ATAC_peaks)
venn_from_gr_list(filtered_2_peak_list, "venn_ChIP_ATAC_filtered_stringent")

# export the peaks into bed files
imap(filtered_2_peak_list[1:2], ~export.bed(.x, 
                                     con=glue("{out}/{.y}_qVal6_FC6_width10k.bed"), 
                                     format = "bed")) 

export.bed(filtered_2_peak_list[[3]], 
         con=glue("{out}/FOXR2_ATAC_qVal4_FC2_width10k.bed"), 
         format = "bed")
```

Get a bed file with intersecting peaks between the n=2 FOXR2 CUT&RUN 
and ATAC, for IGV:

```{r intersect-chip-atac}

# Create a bed file of the overlap between the 3 sets of peaks (ChIP FOXR2 rep 1&2, ATAC FOXR2-p53 rep1) 
overlap_all <- Reduce(function(x, y) subsetByOverlaps(x, y, minoverlap = 10), filtered_2_peak_list)
export.bed(overlap_all, con = glue("{out}/intersect_ChIP_ATAC_qVal6_FC6_width10k.bed"), format = "bed")

# Create the intersect with the exact bases that intersect the ChIP FOXR2 rep 1&2 and ATAC FOXR2-p53 rep1
# Run after: module load bedtools/2.30.0
system(glue("intersectBed -a {out}/rep1_FOXR2_qVal6_FC6_width10k.bed -b {out}/rep2_FOXR2_qVal6_FC6_width10k.bed > {out}/exact_ChIP-intersect_qVal6_FC6_width10k.bed"))
system(glue("intersectBed -a {out}/exact_ChIP-intersect_qVal6_FC6_width10k.bed -b {out}/FOXR2_ATAC_qVal4_FC2_width10k.bed > {out}/exact_intersect_ChIP_ATAC_qVal6_FC6_width10k.bed"))

```

## Peak QC metric distribution

Plot histograms of metrics (signal, p-val, q-val) distributions across
CUT&RUN peaks, with QC thresholds for each metric as a vertical line.

```{r histo-peaks, fig.width = 12, fig.height = 4}

qval_threshold <- c(6, 6, 4)
fold_threshold <- c(6, 6, 2)
width_threshold1 <- 1
width_threshold2 <- 10000

# Q values
# log-log scale
map2(names(peak_list), qval_threshold,
     ~ data.frame(num = peak_list[[.x]]$name,
                  qValue = peak_list[[.x]]$qValue) %>% 
            ggplot(aes(x = log2(qValue))) +
            geom_histogram(bins = 100) +
            xlab("log2(-log10(qValue))") +
            ggtitle(glue("{.x} - q-value")) +
            geom_vline(xintercept = log2(.y), color = "red"))  %>% 
    cowplot::plot_grid(plotlist = ., ncol = 3)

# Fold enrichment
map2(names(peak_list), fold_threshold,
     ~ data.frame(num = peak_list[[.x]]$name,
                  fold_enrichment = peak_list[[.x]]$signalValue) %>% 
            ggplot(aes(x = log2(fold_enrichment))) +
            geom_histogram(bins = 100) +
            xlab("log2(fold_enrichment)") +
            ggtitle(glue("{.x} - signal")) +
            geom_vline(xintercept = log2(.y), color = "red")) %>% 
    cowplot::plot_grid(plotlist = ., ncol = 3)

# Peak width

width_df <- function(gr){
    
    df <- data.frame(num = gr$name)
    df$width <- width(gr)
    return(df)
    
}

imap(peak_list,
     ~ width_df(.x) %>% 
            ggplot(aes(x = width)) +
            geom_histogram(bins = 300) +
            xlab("peak width") +
            ggtitle(glue("{.y} - width")) +
            geom_vline(xintercept = width_threshold1, color = "red") +
            geom_vline(xintercept = width_threshold2, color = "red"))  %>% 
    cowplot::plot_grid(plotlist = ., ncol = 3)


```

# Motif analysis

Motif analysis with HOMER was conducted with scripts in the following directory:

```
code/scripts/ChIP
```

Originally written and run by Steven Hébert.

```{r motif_analysis, cache=F}

# Load Homer motif results
motifs <- read_tsv(glue("{out}/motifs_size100_100/knownResults.txt"))

# Adjust p-values and count the number of significant motifs
motifs_adj <- motifs %>%
  mutate(padj_BH = p.adjust(`P-value`, method = "BH")) 
nrow(filter(motifs_adj, padj_BH < 0.05))



```

Plot the top motifs and their -log10 adjusted p-value, coloring by the motif type
(ETS, SOX, other):

```{r plot-motifs}

motifs_gg <- motifs %>%
  mutate(padj_BH = p.adjust(`P-value`, method = "BH")) %>%
  slice_max(order_by = -`Log P-value`, n = 13, with_ties = T) %>%
  mutate(Motif_name = str_split(`Motif Name`, "/", simplify = T)[,1]) %>% 
  mutate(Motif_group = case_when(grepl(.$Motif_name, pattern = "ETS") ~ "ETS",
                                 grepl(.$Motif_name, pattern = "Sox") ~ "SOX",
                                 TRUE                                        ~ "Other"))

motifs_gg %>% 
  ggplot(aes(x = -log10(padj_BH), y = reorder(Motif_name, -padj_BH))) +
  geom_bar(stat = "identity", aes(fill = Motif_group)) +
  ylab("Motif") +
  xlab("-log10(padj)") +
  scale_fill_manual(values = c("ETS" = "darkgoldenrod1",
                               "SOX" = "darkolivegreen3",
                               "Other" = "grey70")) +
  theme(legend.position = "none")

```

Reorder the bars to group the motif types together, and rank by adjusted p-value
within each of the groups. This is the final plot used in the publication.

```{r plot-motifs-reorder}

df <- motifs_gg %>% 
    arrange(padj_BH) %>% 
    dplyr::select(Motif_name, Motif_group) 

order <- c(df %>% filter(Motif_group == "ETS") %>% .$Motif_name,
           df %>% filter(Motif_group == "SOX") %>% .$Motif_name,
           df %>% filter(Motif_group == "Other") %>% .$Motif_name)

motifs_gg %>% 
  mutate(Motif_name = factor(Motif_name, levels = rev(order))) %>% 
  ggplot(aes(x = -log10(padj_BH), y = Motif_name)) +
  geom_bar(stat = "identity", aes(fill = Motif_group)) +
  ylab("Motif") +
  xlab("-log10(padj)") +
  scale_fill_manual(values = c("ETS" = "darkgoldenrod1",
                               "SOX" = "darkolivegreen3",
                               "Other" = "grey70")) +
  theme(legend.position = "none")


```


<!-- END MATTER, insert reproducibility info -->

```{r footer, echo = FALSE, results = 'asis', warning = FALSE, cache = FALSE}

# Knit child document with header
res <- knitr::knit_child(here("include", "footer.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF END MATTER -->